---
title: "Preregistration: Alienation and Trust in Climate Scientists in the US"

abstract: |
  Trust in climate scientists is declining and deeply polarized in the United States.
  Prominent psychological accounts attribute distrust to cognitive deficiencies — lack of
  knowledge, motivated reasoning, or conspiracist thinking — treating distrust as a product
  of individual irrationality. We test an alternative: the alienation model of science
  distrust, which holds that distrust stems from structural exclusion from science rather
  than individual cognitive failings. People who are structurally excluded from science —
  institutionally, socially, geographically, or informationally — will be more likely to
  feel alienated from science, and therefore more likely to distrust scientists. This distrust
  may be rational: those who are structurally excluded from an institution have less influence
  over its decisions, which increases their vulnerability and may warrant greater wariness.
  We derive and test specific individual-level predictions of the alienation model using data
  from a large, nationally representative US sample (N ≈ 22,000). All analyses are
  correlational, and all hypotheses and analytical procedures are pre-registered prior to
  data collection.

date: "2026-05-10"

format:
  html:
    code-fold: false

execute:
  echo: false
  warning: false
  message: false
  fig-width: 6.5
  fig-height: 4
  fig-dpi: 300
---

```{r packages}
#| echo: false

library(tidyverse)
library(broom)
library(sandwich)
library(lmtest)
library(marginaleffects)
library(tinytable)
library(patchwork)
library(psych)
library(here)
```

```{r functions}
list.files(here("R/functions"), pattern = "\\.R$", full.names = TRUE) |>
  walk(source)
```

```{r load-data}
data <- readRDS(here("data/simulation/simulated_data.rds"))
```

::: callout-note
This preregistration has not yet been registered on the OSF and may change within the next days.
:::

\clearpage

# Background

Trust in climate scientists is a critical determinant of climate change belief and policy support, yet it is declining and deeply polarized in the United States. Much of the existing literature explains distrust in scientists through individual-level psychological mechanisms: people distrust scientists because they lack scientific knowledge, engage in motivated reasoning, or hold conspiracist or populist thinking styles. These accounts share a common assumption — that distrust reflects some form of cognitive failure or irrationality on the part of the distruster.

The *alienation model* of science distrust, developed by Gauchat [-@gauchat2011; -@gauchat2012politicization], offers a structurally grounded alternative. Rooted in the work of theorists including Beck [-@beck1992risk], Giddens [-@giddens1991modernity], and Habermas [-@habermas1989structural], the model argues that increasingly complex societies require increasingly specialized governance expertise, producing technocratic institutions run by knowledge elites. Those who are excluded from these institutions — who lack the cultural capital, social proximity, geographic access, or informational exposure to participate — experience a loss of agency and a sense of alienation. This alienation, the model predicts, rationally translates into distrust: people who have no influence over an institution's decisions are more vulnerable to its choices and have less reason to extend it trust [@desmondVulnerabilityTrust2023].

This framing has important implications. It suggests that distrust in climate scientists among demographically underrepresented groups — racial minorities, people with less education, rural residents — is not primarily a product of irrationality or misinformation, but a predictable response to structural exclusion. Despite its theoretical appeal, the alienation model has not yet been directly tested with respect to climate scientists. Gauchat's own empirical work tested only the high-level prediction that increasing technocratization would produce declining aggregate trust in science. More specific, individual-level predictions derived from the model remain untested.

---

# Theoretical Framework

We define alienation as a lack of access to climate scientists, climate science, and climate science institutions. In this research project, we focus on *perceived alienation*, based on people's self-reports. We distinguish four dimensions of perceived alienation, informed by research on psychological distance to science [@rutjensPsychologicalDistanceScience2022; @veckalovPsychologicalDistanceScience2024]:

**Institutional alienation** refers to the lack of representation within and influence over scientific institutions. People who do not see themselves reflected in the demographics of climate scientists, or who feel they have no voice in how scientific knowledge is produced and used, experience institutional alienation.

**Social alienation** refers to the social distance between oneself and climate scientists as people---a sense that scientists inhabit a different social world, with different values, backgrounds, and life experiences.

**Spatial alienation** refers to geographic distance from scientific institutions. People who live far from universities, research centers, and other sites of scientific activity have fewer opportunities for incidental contact with scientists and their work.

**Informational alienation** refers to limited exposure to scientific information in everyday life---through news, social media, personal conversations, or in-person events. People who rarely encounter scientific information are informationally distant from the scientific world.

We propose that these four dimensions are empirically distinguishable and may have different relationships with trust in climate scientists. We test predictions about each dimension separately.

# Hypotheses and Research Questions

## Hypotheses

**H1 — Association between alienation and trust:** Perceptions of alienation are negatively associated with trust in climate scientists. 

**H2 — Epistemic autonomy as moderator of the alienation-trust relationship:** Climate scientists claim epistemic authority over matters of public concern. People with a high need for epistemic autonomy, who are motivated to form their own judgments rather than defer to authorities, should be particularly distrusting of climate scientists when they experience alienation. We therefore predict that need for epistemic autonomy moderates the relationship between alienation and distrust, such that the alienation–distrust association is stronger among high-autonomy individuals.

For both hypotheses, we test the four dimensions of alienation separately. 

## Research Questions

**RQ1 — Structural predictors of institutional and social alienation:** Do people who are demographically underrepresented among climate scientists---by race, education, gender, or social class---report higher institutional and social alienation? This tests whether self-reported perceptions of alienation map onto objective indicators of structural underrepresentation.

**RQ2 — Structural predictors of spatial alienation:** Do people who live further from scientific institutions---operationalized as geographic distance from the nearest research university---report higher spatial alienation? This tests whether geographic distance predicts perceived distance.

**RQ3 — Alienation and dimensions of trust:** Alienated people may still recognize climate scientists' technical competence, but have more reason to doubt their benevolence, integrity, and openness---dimensions of trust that reflect whether an institution acts in one's interest rather than merely whether it is capable. We therefore ask whether alienation is more strongly associated with distrust along the dimensions of benevolence, integrity, and openness than along the dimension of competence.

**RQ4 — Alienation and policy role:** People who are alienated from climate scientists may be more resistant to granting scientists a role in shaping public policy. We ask whether alienation predicts opposition to scientists having influence over climate policy.

# Data

## Source

We use pre-treatment data from the **Strengthening trust in climate scientists megastudy** (N ≈ 22,000), a large collaborative experiment testing 20 interventions to strengthen trust in climate scientists in the United States. Participants are recruited from a national, non-probability opt-in panel (CloudResearch) with cross-quotas on gender × age and gender × race based on 2024 US Census Bureau population estimates.

This project presents correlational analyses of the megastudy pre-treatment measures. No experimental comparisons are made. **No data has been collected yet**. 

## Ethics

The megastudy has received ethics approval from Stanford University (IRB-85756) and ETH Zurich (26 ETHICS-093).

# Measures

## Alienation from climate science

Except for informational alienation, all items use a 7-point agree/disagree response scale (1 = *strongly disagree*, 7 = *strongly agree*). Higher scores indicate greater alienation. We will use composite scores (the mean) of all items present in one dimension. 

### Institutional alienation (2 items)

- *The prospect of working as a climate scientist has always seemed beyond my reach.*
- *Careers in climate research are accessible only to a privileged few.*

### Social alienation (2 items)

- *Climate scientists have a different social background than me.*
- *Climate scientists move in different social circles than me.*

### Spatial alienation (2 items)

- *Climate science has no positive impact on my local area.*
- *Very few climate scientists live or work in my local area.*

### Informational alienation (6 items)

"How often do you see or hear information about climate change in the following places?"

1. Traditional media (e.g., newspapers, TV, radio)
2. Online news (e.g., news websites, podcasts, YouTube)
3. Social media (e.g., Facebook, TikTok, Instagram)
4. Fiction (e.g., films, series, books, comics)
5. Personal conversations (e.g., talking with friends or family, text messages, messaging apps)
6. In-person events (e.g., museums, public talks)

Response options: 1 = *Never*, 2 = *Rarely*, 3 = *Occasionally*, 4 = *Frequently*, 5 = *Very frequently*. Items are **reverse-coded** before computing the composite, so that higher scores indicate less exposure (more informational alienation).

## Trust in climate scientists

### Single-item trust

The primary outcome is a single outcome trust measure, assessed pre-treatment in the megastudy.

"How much do you trust climate scientists?" (slider, 0–100).

## Multidimensional trust (secondary outcome, 12 items)

For a robustness check, we use a more detailed 12-item trust scale covering four subdimensions (3 items each), assessed on 0–100 sliders (0 = most negative, 100 = most positive); see @tbl-trust-scale. Subdimension composites are computed as the mean of the three items within each subdimension. The overall composite (`trust_multidimensional`) is the mean of all 12 items. 

The reason this scale is used only for a robustness check is that, in the megastudy, it is asked "post-treatment". We can therefore only use answers from the control group for unconfounded results.

```{r}
#| label: tbl-trust-scale
#| tbl-cap: "Multidimensional trust scale: subdimensions and bipolar item anchors. Each item is rated on a 0–100 slider."

library(tinytable)

tibble::tribble(
  ~Subdimension, ~`Items (bipolar anchors)`,
  "Competence",  "incompetent–competent; unintelligent–intelligent; unqualified–qualified",
  "Integrity",   "dishonest–honest; unethical–ethical; insincere–sincere",
  "Benevolence", "unconcerned–concerned; uneager–eager to improve lives; inconsiderate–considerate",
  "Openness",    "not open–open to feedback; unwilling–willing to be transparent; little–great attention to others' views"
) |>
  tt(width = c(0.3, 0.7)) 
```

## Need for epistemic autonomy (6 items)

[@beebePitfallsEpistemicAutonomy2024; adapted]

"Please indicate how much you agree or disagree with the following statements" (1 = *strongly disagree*, 7 = *strongly agree*):

1. *I like to think things through for myself.*
2. *I like to figure things out for myself.*
3. *I like to make up my own mind about things.*
4. *I only believe something if I can see for myself that it is true.*
5. *I don't go along with the opinions of others without thinking things through for myself.*
6. *I have never really questioned the things I have been taught to believe.* [reverse-scored]

We will use a composite measure (mean) of all items items, reverse-coding item 6.

## Structural indicators (for RQ1 and RQ2)

- **Race/ethnicity**: White, Black, Hispanic, Asian, Other (categorical)
- **Education**: 6-level ordinal (less than high school → doctorate)
- **Gender**: Male, Female, Other
- **Income**: 5-level ordinal (\<\$30k to ≥\$168k)
- **Geographic distance** (`dist_inst_km`): distance (km) from the participant's ZIP code to the nearest scientific institution (R1/R2 research university or major federal research lab), computed using ZIP-code centroid coordinates

## Policy role of scientists (for RQ4)

Four items assessing support for scientists having a role in climate policy (e.g., collaborating with policymakers, advocating for policies, communicating findings; 0–100 sliders). We will use a composite score (mean) of all items.

---

# Analysis Plan

## Overview

All models are estimated using **OLS regression with heteroskedasticity-robust standard errors** (HC2, via the `sandwich` package). All continuous predictors and outcomes are **standardized** (z-scored) before model fitting, so reported coefficients are standardized betas and directly comparable across dimensions. Within each hypothesis and research question, multiple comparisons are corrected using the **Benjamini-Hochberg (BH) false discovery rate procedure** [@benjaminiControllingFalseDiscovery1995].

Code chunks below demonstrate each analysis on simulated data. Function definitions are printed before their first use so the exact model specification is part of the preregistration record.

## H1: Association between alienation and trust

For each of the four alienation dimensions, we run separate regressions. We regress the single-item pre-treatment trust measure (`trust_pre`) on the respective alienation composite. We do not include covariates. BH correction is applied across the four tests (one per dimension).

```{r h1-function}
#| results: asis
print_a_function_from_file("run_ols_model")
```

```{r h1-models}
#| echo: true

alienation_dims <- c(
  "alien_inst_mean", "alien_social_mean",
  "alien_spatial_mean", "alien_info_mean"
)

h1_results <- map_dfr(alienation_dims, \(pred)
  run_ols_model(data, outcome = "trust_pre", predictor = pred)
) |>
  mutate(p_adjusted = p.adjust(p.value, method = "BH"))

h1_results |>
  select(predictor, estimate, std.error, statistic, p.value, p_adjusted, n) |>
  mutate(predictor = str_remove(predictor, "_mean") |>
           str_remove("alien_") |> str_to_title()) |>
  rounded_numbers() |>
  rename(
    Dimension = predictor, Beta = estimate, SE = std.error,
    t = statistic, p = p.value, `p (BH-adj)` = p_adjusted, N = n
  ) |>
  tt(caption = "H1: Effect of alienation dimensions on single-item trust (standardized beta, HC2 SEs)")
```

```{r h1-plot}
#| fig-cap: "H1: Standardized associations between alienation dimensions and single-item trust (trust_pre). Colored points: BH-adjusted p < .05."

h1_results |>
  mutate(predictor = recode(predictor,
    alien_inst_mean    = "Institutional",
    alien_social_mean  = "Social",
    alien_spatial_mean = "Spatial",
    alien_info_mean    = "Informational"
  )) |>
  plot_coefficients(label_col = "predictor",
                    title     = "H1: Alienation and Trust in Climate Scientists")
```

## H2: Epistemic autonomy as moderator

For each alienation dimension, we estimate an interaction model:

$$
\text{trust\_pre} \sim \text{alienation}_d \times \text{epist\_auton\_mean} + \text{age} + \text{gender} + \text{race}
$$

We report the interaction coefficient (alienation × epistemic autonomy) for each dimension, with BH correction across the four interaction tests.

```{r h2-function}
#| results: asis
print_a_function_from_file("run_interaction_model")
```

```{r h2-models}
#| echo: true

h2_results <- map_dfr(alienation_dims, \(pred)
  run_interaction_model(
    data,
    outcome    = "trust_pre",
    predictor  = pred,
    moderator  = "epist_auton_mean",
    covariates = c("age", "gender", "race")
  )
) |>
  filter(str_detect(term, ":")) |>
  mutate(p_adjusted = p.adjust(p.value, method = "BH"))

h2_results |>
  select(predictor, moderator, estimate, std.error, statistic, p.value, p_adjusted, n) |>
  mutate(predictor = str_remove(predictor, "_mean") |>
           str_remove("alien_") |> str_to_title()) |>
  rounded_numbers() |>
  rename(
    `Alienation dim.` = predictor, Moderator = moderator,
    Beta = estimate, SE = std.error, t = statistic,
    p = p.value, `p (BH-adj)` = p_adjusted, N = n
  ) |>
  tt(caption = "H2: Interaction of each alienation dimension with epistemic autonomy predicting trust_pre")
```

```{r h2-plot}
#| fig-cap: "H2 (example: institutional alienation): Predicted trust across epistemic autonomy levels for low vs. high institutional alienation."

plot_interaction(
  data,
  outcome    = "trust_pre",
  predictor  = "alien_inst_mean",
  moderator  = "epist_auton_mean",
  pred_label = "Institutional alienation",
  mod_label  = "Epistemic autonomy",
  out_label  = "Trust (pre-treatment)"
)
```

## RQ1: Structural predictors of institutional and social alienation

We regress each alienation outcome (`alien_inst_mean`, `alien_social_mean`) separately on each structural predictor: race (White as reference), education, gender (Male as reference), and social class (Middle class as reference). Each predictor is entered in a separate model that adjusts for age and the remaining demographic covariates. BH correction is applied within each alienation outcome across all structural predictors.

```{r rq1-models}
#| echo: true

structural_preds <- c("race", "education", "gender", "social_class")
rq1_outcomes     <- c("alien_inst_mean", "alien_social_mean")

rq1_results <- map_dfr(rq1_outcomes, \(out)
  map_dfr(structural_preds, \(pred)
    run_ols_model(
      data,
      outcome    = out,
      predictor  = pred,
      covariates = c("age", setdiff(structural_preds, pred))
    )
  ) |>
    mutate(p_adjusted = p.adjust(p.value, method = "BH"),
           outcome    = out)
)

rq1_results |>
  select(outcome, predictor, term, estimate, std.error, p.value, p_adjusted) |>
  mutate(
    outcome   = str_remove(outcome, "_mean") |> str_remove("alien_") |> str_to_title(),
    predictor = str_replace_all(predictor, "_", " ") |> str_to_title(),
    term      = str_replace_all(term, "_", " ")
  ) |>
  rounded_numbers() |>
  rename(
    Outcome = outcome, Variable = predictor, Level = term,
    Beta = estimate, SE = std.error, p = p.value, `p (BH-adj)` = p_adjusted
  ) |>
  tt(caption = "RQ1: Structural predictors of institutional and social alienation")
```

## RQ2: Structural predictor of spatial alienation

We regress spatial alienation on log-transformed geographic distance to the nearest research university, adjusting for age, gender, race, and education. This is a single test; no BH correction needed.

```{r rq2-model}
#| echo: true

rq2_result <- run_ols_model(
  data |> mutate(log_dist_inst_km = log(dist_inst_km)),
  outcome    = "alien_spatial_mean",
  predictor  = "log_dist_inst_km",
  covariates = c("age", "gender", "race", "education")
)

rq2_result |>
  select(predictor, estimate, std.error, statistic, p.value, n) |>
  mutate(predictor = "log(distance to university, km)") |>
  rounded_numbers() |>
  rename(Predictor = predictor, Beta = estimate, SE = std.error,
         t = statistic, p = p.value, N = n) |>
  tt(caption = "RQ2: Geographic distance predicting spatial alienation")
```

## RQ3: Alienation and dimensions of trust

The multidimensional trust scale is measured post-treatment in the megastudy. To obtain unconfounded estimates, we restrict this analysis to **control group participants only**. For each of the four alienation dimensions, we regress each trust subdimension separately and apply BH correction across the four subdimension tests within each alienation dimension.

```{r rq3-models}
#| echo: true

trust_dims  <- c("trust_competence", "trust_integrity",
                 "trust_benevolence", "trust_openness")
data_control <- data |> filter(condition == "control")

rq3_results <- map_dfr(alienation_dims, \(pred)
  map_dfr(trust_dims, \(out)
    run_ols_model(data_control, outcome = out, predictor = pred)
  ) |>
    mutate(p_adjusted = p.adjust(p.value, method = "BH"))
)

rq3_results |>
  filter(predictor == "alien_inst_mean") |>    # example: institutional alienation
  select(outcome, estimate, std.error, p.value, p_adjusted) |>
  mutate(outcome = str_remove(outcome, "trust_") |> str_to_title()) |>
  rounded_numbers() |>
  rename(Dimension = outcome, Beta = estimate, SE = std.error,
         p = p.value, `p (BH-adj)` = p_adjusted) |>
  tt(caption = "RQ3 (example: institutional alienation): Effect on each trust subdimension (control group only)")
```

```{r rq3-plot}
#| fig-cap: "RQ3 (example: institutional alienation): Associations with each trust subdimension, control group only."

rq3_results |>
  filter(predictor == "alien_inst_mean") |>
  mutate(outcome = recode(outcome,
    trust_competence  = "Competence",
    trust_integrity   = "Integrity",
    trust_benevolence = "Benevolence",
    trust_openness    = "Openness"
  )) |>
  plot_coefficients(label_col = "outcome",
                    title = "RQ3: Institutional alienation and trust subdimensions")
```

## RQ4: Alienation and policy role

We regress policy role support (`policy_role_mean`) on each alienation dimension separately, without covariates (mirroring H1). BH correction across four tests.

```{r rq4-models}
#| echo: true

rq4_results <- map_dfr(alienation_dims, \(pred)
  run_ols_model(data, outcome = "policy_role_mean", predictor = pred)
) |>
  mutate(p_adjusted = p.adjust(p.value, method = "BH"))

rq4_results |>
  select(predictor, estimate, std.error, p.value, p_adjusted, n) |>
  mutate(predictor = str_remove(predictor, "_mean") |>
           str_remove("alien_") |> str_to_title()) |>
  rounded_numbers() |>
  rename(
    Dimension = predictor, Beta = estimate, SE = std.error,
    p = p.value, `p (BH-adj)` = p_adjusted, N = n
  ) |>
  tt(caption = "RQ4: Alienation dimensions predicting policy role support")
```

```{r rq4-plot}
#| fig-cap: "RQ4: Alienation dimensions and support for scientists' policy role."

rq4_results |>
  mutate(predictor = recode(predictor,
    alien_inst_mean    = "Institutional",
    alien_social_mean  = "Social",
    alien_spatial_mean = "Spatial",
    alien_info_mean    = "Informational"
  )) |>
  plot_coefficients(label_col = "predictor",
                    title = "RQ4: Alienation and Policy Role Support")
```

---

# Multiple Comparisons

We apply BH-FDR correction **within each hypothesis and research question separately** (not across all tests in the study). The correction sets apply to:

- **H1**: 4 tests (one per alienation dimension; outcome = `trust_pre`)
- **H2**: 4 tests (one interaction per alienation dimension; outcome = `trust_pre`)
- **RQ1**: tests within each alienation outcome (institutional, social), across structural predictors
- **RQ2**: 1 test — no correction needed
- **RQ3**: 4 tests per alienation dimension (one per trust subdimension; control group only)
- **RQ4**: 4 tests (one per alienation dimension; outcome = `policy_role_mean`)

---

# Supplementary Materials

## Scale Reliability

```{r reliability}
#| echo: true

get_alpha <- function(df, cols) {
  psych::alpha(df[, cols], check.keys = FALSE)$total$raw_alpha
}

tibble(
  Scale = c(
    "Institutional alienation",
    "Social alienation",
    "Spatial alienation",
    "Informational alienation",
    "Trust: Competence",
    "Trust: Integrity",
    "Trust: Benevolence",
    "Trust: Openness",
    "Trust: Multidimensional (all 12 items)",
    "Epistemic autonomy"
  ),
  Items = c(2, 2, 2, 6, 3, 3, 3, 3, 12, 6),
  Alpha = c(
    get_alpha(data, c("alien_inst_1", "alien_inst_2")),
    get_alpha(data, c("alien_social_1", "alien_social_2")),
    get_alpha(data, c("alien_spatial_1", "alien_spatial_2")),
    get_alpha(data, paste0("alien_info_", 1:6, "_r")),
    get_alpha(data, paste0("trust_competence_", 1:3)),
    get_alpha(data, paste0("trust_integrity_", 1:3)),
    get_alpha(data, paste0("trust_benevolence_", 1:3)),
    get_alpha(data, paste0("trust_openness_", 1:3)),
    get_alpha(data, c(paste0("trust_competence_", 1:3),
                      paste0("trust_integrity_", 1:3),
                      paste0("trust_benevolence_", 1:3),
                      paste0("trust_openness_", 1:3))),
    get_alpha(data, c("epist_auton_1", "epist_auton_2", "epist_auton_3",
                      "epist_auton_4", "epist_auton_5", "epist_auton_6r"))
  )
) |>
  mutate(Alpha = round(Alpha, 2)) |>
  tt(caption = "Internal consistency (Cronbach's alpha) of all multi-item scales (simulated data)")
```

## Simulated sample characteristics

```{r sample-characteristics}
#| echo: true

data |>
  select(age, gender, race, education, income, social_class, urban_rural) |>
  summary() |>
  knitr::kable(caption = "Simulated sample characteristics (N = 500)")
```

# References
