---
title: "Scientific Institutions: Location Data & ZIP-Code Distance Pipeline"
author: "Jan Pfänder"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This document prepares the `dist_inst_km` variable used in the alienation–trust
preregistration: the geodesic distance (km) from each participant's ZIP code to
the nearest scientific institution.

**What counts as a scientific institution?**

We include two categories:

1. **Research universities** — Carnegie R1 (Doctoral: Very High Research
   Activity) and R2 (Doctoral: High Research Activity) institutions, sourced
   from the Integrated Postsecondary Education Data System (IPEDS) via the
   Urban Institute's `educationdata` package.

2. **Major federal research labs** — approximately 50 facilities operated by
   DOE, NIH, NIST, NOAA, NASA, and USGS, with coordinates sourced from each
   agency's website and verified against OpenStreetMap.

**Why log-transform?** Geographic distance is right-skewed (a few very remote
participants drive the distribution). All analyses use `log(dist_inst_km)`.

**Outputs** (saved to `data/institutions/`, gitignored):

- `institutions.rds` — clean institution dataset with coordinates
- `zipcode_distances.rds` — ZIP-code lookup table: one row per ZIP, with
  nearest institution and distance in km

---

## Setup

```{r packages}
library(tidyverse)
library(here)
library(sf)
library(geosphere)
library(zipcodeR)
```

```{r create-dir}
dir.create(here("data", "institutions"), recursive = TRUE, showWarnings = FALSE)
```

---

## Part 1: Research Universities (IPEDS)

We download the IPEDS Header/Directory file (HD2022) directly from NCES.
This file is the authoritative source for institution characteristics including
coordinates and Carnegie classification codes. Downloading directly avoids any
API dependency.

Carnegie classification codes of interest (variable `C18BASIC` in IPEDS):

| Code | Label |
|------|-------|
| 15   | Doctoral Universities: Very High Research Activity (R1) |
| 16   | Doctoral Universities: High Research Activity (R2) |

```{r ipeds-download}
#| cache: true

# Download IPEDS HD2022 (Header/Directory) — ~3 MB zip, ~6,000 institutions.
# Cached by knitr after first run.
ipeds_zip  <- tempfile(fileext = ".zip")
ipeds_dir  <- tempdir()

download.file(
  url      = "https://nces.ed.gov/ipeds/datacenter/data/HD2022.zip",
  destfile = ipeds_zip,
  quiet    = TRUE,
  mode     = "wb"
)

unzip(ipeds_zip, exdir = ipeds_dir)

# The main CSV is lowercase in the zip; find it regardless of case
csv_file   <- list.files(ipeds_dir, pattern = "(?i)^hd2022\\.csv$",
                         full.names = TRUE)[1]
ipeds_raw  <- read_csv(csv_file, show_col_types = FALSE)

# Key columns: INSTNM, CITY, STABBR, LATITUDE, LONGITUD, C18BASIC
glimpse(ipeds_raw |> select(INSTNM, CITY, STABBR, LATITUDE, LONGITUD, C18BASIC))
```

```{r ipeds-clean}
universities <- ipeds_raw |>
  filter(
    C18BASIC %in% c(15, 16),        # R1 and R2 only
    !is.na(LATITUDE),
    !is.na(LONGITUD),
    # Drop non-contiguous US (Alaska, Hawaii, territories)
    LATITUDE  > 24 & LATITUDE  < 50,
    LONGITUD  > -125 & LONGITUD < -66
  ) |>
  transmute(
    name             = INSTNM,
    city             = CITY,
    state_abbr       = STABBR,
    lat              = LATITUDE,
    lon              = LONGITUD,
    carnegie_class   = if_else(C18BASIC == 15, "R1", "R2"),
    institution_type = "university"
  )

cat("R1 universities:", sum(universities$carnegie_class == "R1"), "\n")
cat("R2 universities:", sum(universities$carnegie_class == "R2"), "\n")
cat("Total:          ", nrow(universities), "\n")
```

```{r universities-map}
universities_sf <- st_as_sf(universities, coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = universities_sf,
          aes(color = carnegie_class), size = 1.2, alpha = 0.7) +
  scale_color_manual(values = c(R1 = "#2c7bb6", R2 = "#abd9e9"),
                     name = "Carnegie") +
  labs(title = "Research universities included (R1 + R2)",
       caption = "Source: IPEDS 2022 via Urban Institute educationdata package") +
  theme_void() +
  theme(legend.position = "bottom")
```

---

## Part 2: Federal Research Labs

Manually curated list of ~50 major federal research facilities. Coordinates
verified against agency websites and OpenStreetMap. This list intentionally
covers the largest, most prominent labs — it is not exhaustive of every federal
facility, but captures the institutions most relevant to structural proximity
to organized science.

```{r federal-labs}
federal_labs <- tribble(
  ~name,                                          ~agency,  ~lat,     ~lon,
  # ── Department of Energy national labs ─────────────────────────────────
  "Argonne National Laboratory",                  "DOE",   41.7059, -87.9837,
  "Brookhaven National Laboratory",               "DOE",   40.8674, -72.8776,
  "Fermi National Accelerator Laboratory",        "DOE",   41.8319, -88.2617,
  "Idaho National Laboratory",                    "DOE",   43.5246,-112.9541,
  "Lawrence Berkeley National Laboratory",        "DOE",   37.8755,-122.2477,
  "Lawrence Livermore National Laboratory",       "DOE",   37.6868,-121.7051,
  "Los Alamos National Laboratory",               "DOE",   35.8800,-106.2951,
  "National Renewable Energy Laboratory",         "DOE",   39.7405,-105.1686,
  "Oak Ridge National Laboratory",                "DOE",   35.9310, -84.3100,
  "Pacific Northwest National Laboratory",        "DOE",   46.3440,-119.2785,
  "Sandia National Laboratories (NM)",            "DOE",   34.9653,-106.3969,
  "Sandia National Laboratories (CA)",            "DOE",   37.9717,-121.7046,
  "SLAC National Accelerator Laboratory",         "DOE",   37.4160,-122.2034,
  "Ames National Laboratory",                     "DOE",   42.0267, -93.7317,
  "Jefferson Lab",                                "DOE",   37.0977, -76.4791,
  "Savannah River National Laboratory",           "DOE",   33.3487, -81.7229,
  "National Energy Technology Laboratory",        "DOE",   39.9279, -79.9701,
  # ── National Institutes of Health ──────────────────────────────────────
  "NIH Main Campus",                              "NIH",   39.0007, -77.1055,
  "NIH Rocky Mountain Labs",                      "NIH",   46.8574,-113.9867,
  "NIH Research Triangle Park",                   "NIH",   35.8973, -78.8622,
  # ── National Institute of Standards and Technology ──────────────────────
  "NIST Gaithersburg",                            "NIST",  39.1330, -77.2177,
  "NIST Boulder",                                 "NIST",  40.0050,-105.2625,
  # ── NOAA research labs ──────────────────────────────────────────────────
  "NOAA Earth System Research Lab (Boulder)",     "NOAA",  39.9950,-105.2639,
  "NOAA Geophysical Fluid Dynamics Lab",          "NOAA",  40.3502, -74.6573,
  "NOAA Atlantic Oceanographic Lab (Miami)",      "NOAA",  25.7306, -80.1625,
  "NOAA Pacific Marine Environmental Lab",        "NOAA",  47.6456,-122.3130,
  "NOAA National Severe Storms Lab",              "NOAA",  35.1817, -97.4432,
  "NOAA Great Lakes Environmental Research Lab",  "NOAA",  42.2986, -83.7082,
  # ── NASA field centers ───────────────────────────────────────────────────
  "NASA Ames Research Center",                    "NASA",  37.4094,-122.0640,
  "NASA Armstrong Flight Research Center",        "NASA",  34.9594,-117.8822,
  "NASA Glenn Research Center",                   "NASA",  41.4273, -81.8633,
  "NASA Goddard Space Flight Center",             "NASA",  38.9915, -76.8404,
  "NASA Jet Propulsion Laboratory",               "NASA",  34.2014,-118.1720,
  "NASA Johnson Space Center",                    "NASA",  29.5593, -95.0938,
  "NASA Kennedy Space Center",                    "NASA",  28.5729, -80.6490,
  "NASA Langley Research Center",                 "NASA",  37.0837, -76.3694,
  "NASA Marshall Space Flight Center",            "NASA",  34.7293, -86.6650,
  "NASA Stennis Space Center",                    "NASA",  30.3629, -89.6003,
  # ── USGS science centers ─────────────────────────────────────────────────
  "USGS National Center (Reston)",                "USGS",  38.9476, -77.3673,
  "USGS Western Region (Menlo Park)",             "USGS",  37.4574,-122.1730,
  "USGS Central Region (Denver)",                 "USGS",  39.7543,-104.8797,
  "USGS Woods Hole Coastal Science Center",       "USGS",  41.5250, -70.6710,
  "USGS Columbia Environmental Research Center",  "USGS",  38.9387, -92.3341,
  "USGS Patuxent Wildlife Research Center",       "USGS",  38.9980, -76.8155,
  # ── CDC / EPA ────────────────────────────────────────────────────────────
  "CDC / ATSDR (Atlanta)",                        "CDC",   33.7993, -84.3277,
  "EPA Office of Research & Development (RTP)",   "EPA",   35.8905, -78.8751,
  "EPA National Risk Management Lab (Cincinnati)","EPA",   39.1290, -84.5122
) |>
  mutate(institution_type = "federal_lab",
         carnegie_class   = NA_character_,
         city             = NA_character_,
         state_abbr       = NA_character_)

cat("Federal labs:", nrow(federal_labs), "\n")
```

```{r labs-map}
labs_sf <- st_as_sf(federal_labs, coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = labs_sf, aes(color = agency), size = 2) +
  labs(title = "Federal research labs included",
       color = "Agency",
       caption = "Coordinates verified against agency websites and OpenStreetMap") +
  theme_void() +
  theme(legend.position = "bottom")
```

---

## Part 3: Combined Institution Dataset

```{r combine}
institutions <- bind_rows(
  universities |> select(name, city, state_abbr, lat, lon,
                          carnegie_class, institution_type),
  federal_labs |> select(name, city, state_abbr, lat, lon,
                          carnegie_class, institution_type)
)

institutions |>
  count(institution_type, carnegie_class) |>
  knitr::kable(caption = "Institution counts by type")
```

```{r combined-map}
inst_sf <- st_as_sf(institutions, coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = inst_sf,
          aes(color = institution_type, size = institution_type),
          alpha = 0.7) +
  scale_color_manual(values = c(university = "#2c7bb6", federal_lab = "#d7191c"),
                     name = NULL) +
  scale_size_manual(values  = c(university = 0.8, federal_lab = 2),
                    guide   = "none") +
  labs(title    = "All scientific institutions",
       subtitle = paste0(sum(institutions$institution_type == "university"),
                         " R1/R2 universities + ",
                         sum(institutions$institution_type == "federal_lab"),
                         " federal labs")) +
  theme_void() +
  theme(legend.position = "bottom")
```

```{r save-institutions}
saveRDS(institutions, here("data", "institutions", "institutions.rds"))
cat("Saved institutions.rds:", nrow(institutions), "rows\n")
```

---

## Part 4: ZIP Code Distance Pipeline

For each US ZIP code, we compute the geodesic distance (km) to the nearest
institution using the Haversine formula. ZIP-code centroids come from
`zipcodeR::zip_code_db`, which covers all ~40,000 US ZIP codes with no API
calls required.

We restrict to the continental US (latitude 24°–50°, longitude −125° to −66°)
to match the participant sample and the institution coverage above.

```{r zip-centroids}
zips <- zip_code_db |>
  filter(
    !is.na(lat), !is.na(lng),
    lat  > 24 & lat  < 50,
    lng > -125 & lng < -66
  ) |>
  select(zip_code = zipcode, zip_lat = lat, zip_lon = lng)

cat("ZIP codes in continental US:", nrow(zips), "\n")
```

```{r nearest-institution}
# Convert to sf for fast nearest-neighbor search
zips_sf  <- st_as_sf(zips, coords = c("zip_lon", "zip_lat"), crs = 4326)
inst_sf2 <- st_as_sf(institutions, coords = c("lon", "lat"),  crs = 4326)

# Find index of nearest institution for each ZIP code
nearest_idx <- st_nearest_feature(zips_sf, inst_sf2)

# Retrieve institution metadata for each matched row
nearest_inst <- institutions[nearest_idx, ] |>
  select(nearest_institution = name, institution_type)

# Compute Haversine distance (km) between each ZIP centroid and its nearest institution
inst_coords <- cbind(institutions$lon[nearest_idx],
                     institutions$lat[nearest_idx])
zip_coords  <- cbind(zips$zip_lon, zips$zip_lat)

dist_km <- distHaversine(zip_coords, inst_coords) / 1000

zipcode_distances <- bind_cols(zips, nearest_inst) |>
  mutate(dist_inst_km = dist_km)

head(zipcode_distances)
```

### Validation

```{r validation}
summary(zipcode_distances$dist_inst_km)
```

```{r spot-checks}
# NYC (10001) → should be near 0; rural Montana (59931) → should be large
zipcode_distances |>
  filter(zip_code %in% c("10001", "59931", "90210", "60601", "73301")) |>
  select(zip_code, nearest_institution, institution_type, dist_inst_km) |>
  mutate(dist_inst_km = round(dist_inst_km, 1)) |>
  knitr::kable(caption = "Spot checks: selected ZIP codes")
```

```{r dist-distribution}
ggplot(zipcode_distances, aes(x = dist_inst_km)) +
  geom_histogram(bins = 80, fill = "#2c7bb6", color = "white", linewidth = 0.2) +
  scale_x_log10(labels = scales::comma) +
  labs(
    x        = "Distance to nearest institution (km, log scale)",
    y        = "Number of ZIP codes",
    title    = "Distribution of dist_inst_km across continental US ZIP codes",
    subtitle = "Right-skewed → log-transform used in analysis"
  ) +
  theme_minimal(base_size = 12)
```

```{r save-distances}
saveRDS(zipcode_distances, here("data", "institutions", "zipcode_distances.rds"))
cat("Saved zipcode_distances.rds:", nrow(zipcode_distances), "rows\n")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
