---
title: "Scientific Institutions: Location Data & ZIP-Code Distance Pipeline"
author: "Jan Pfänder"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This document prepares the proximity-to-science variables used in the
alienation–trust preregistration. The primary measure is `dist_inst_km`, the
geodesic distance (km) from each participant's ZIP code to the nearest
postsecondary institution. We also compute several robustness variants
(per-category nearest distances and counts within 50 km) so the analysis can
test whether the effect is driven by any particular institution type.

**What counts as an institution?**

We include all US postsecondary institutions in the IPEDS Header/Directory
file (HD2022) that grant degrees at the 2-year level or higher
(`ICLEVEL` ∈ {1, 2}). Trade and cosmetology schools (`ICLEVEL == 3`,
less-than-2-year) are excluded, as we do not consider them to be scientific institutions.

Institutions are classified into four mutually exclusive categories using
the 2018 Carnegie basic classification (`C18BASIC`) and `ICLEVEL`:

| `category` | Definition | Approx. N |
|---|---|---|
| `R1`              | Carnegie 15 — Doctoral, Very High Research | ~130 |
| `R2`              | Carnegie 16 — Doctoral, High Research | ~130 |
| `Other 4-year`    | `ICLEVEL == 1` and `C18BASIC ∉ {15, 16}` | ~2,600 |
| `Community college` | `ICLEVEL == 2` | ~1,600 |

**Why log-transform `dist_inst_km`?** Geographic distance is right-skewed (a
few very remote participants drive the distribution). All analyses use
`log(dist_inst_km)`.

**Outputs** (saved to `data/institutions/`, gitignored):

- `institutions.rds` — clean institution dataset with coordinates and `category`
- `zipcode_distances.rds` — one row per continental-US ZIP code with:
  - `dist_inst_km` — distance (km) to nearest institution (primary measure)
  - `nearest_institution`, `nearest_category` — name and category of nearest
  - `dist_r1_km`, `dist_r2_km`, `dist_other4_km`, `dist_cc_km` — per-category nearest
  - `n_within_50km` — overall count within 50 km
  - `n_r1_within_50km`, `n_r2_within_50km`, `n_other4_within_50km`, `n_cc_within_50km`

**Raw data** — must exist locally before rendering this document. Produced by
`R/download_institutions.R`; run once and the file sits in `data/institutions/`
for re-use.

| Source | Script | Output |
|---|---|---|
| IPEDS HD2022 (institution directory) | `R/download_institutions.R` | `data/institutions/institutions_raw.rds` |

---

## Setup

```{r packages}
library(tidyverse)
library(here)
library(sf)
library(geosphere)
library(zipcodeR)
```

## Part 1: Load and categorize HD2022

We read the IPEDS Header/Directory file (HD2022) from the local cache produced
by `R/download_institutions.R`. The 2018 Carnegie basic classification
(`C18BASIC`) distinguishes R1 (code 15) and R2 (code 16); `ICLEVEL` separates
4-year-or-higher (1), 2-year (2), and less-than-2-year (3) institutions.

```{r ipeds-load}
ipeds_path <- here("data", "institutions", "institutions_raw.rds")
if (!file.exists(ipeds_path)) {
  stop("IPEDS HD2022 not found. Run: Rscript R/download_institutions.R")
}
ipeds_raw <- readRDS(ipeds_path)

# Key columns used below: INSTNM, CITY, STABBR, LATITUDE, LONGITUD,
# C18BASIC (Carnegie 2018 basic), ICLEVEL (1=4-yr, 2=2-yr, 3=<2-yr)
glimpse(ipeds_raw |> select(INSTNM, CITY, STABBR, LATITUDE, LONGITUD,
                             C18BASIC, ICLEVEL))
```

```{r ipeds-clean}
institutions <- ipeds_raw |>
  filter(
    ICLEVEL %in% c(1, 2),           # Drop less-than-2-year (trade/cosmetology)
    !is.na(LATITUDE), !is.na(LONGITUD),
    # Continental US only — matches participant sample and the analysis frame
    LATITUDE  > 24 & LATITUDE  < 50,
    LONGITUD  > -125 & LONGITUD < -66
  ) |>
  transmute(
    name       = INSTNM,
    city       = CITY,
    state_abbr = STABBR,
    lat        = LATITUDE,
    lon        = LONGITUD,
    c18basic   = C18BASIC,
    iclevel    = ICLEVEL,
    category   = case_when(
      C18BASIC == 15            ~ "R1",
      C18BASIC == 16            ~ "R2",
      ICLEVEL  == 1             ~ "Other 4-year",
      ICLEVEL  == 2             ~ "Community college"
    )
  ) |>
  mutate(category = factor(category,
    levels = c("R1", "R2", "Other 4-year", "Community college")))

institutions |>
  count(category) |>
  knitr::kable(caption = "Institution counts by category (continental US)")
```

## Part 2: Map of institutions

```{r institutions-map}
#| fig-width: 9
#| fig-height: 5

inst_sf <- st_as_sf(institutions, coords = c("lon", "lat"), crs = 4326)

cat_colors <- c(
  "R1"                = "#2c7bb6",
  "R2"                = "#abd9e9",
  "Other 4-year"      = "#fdae61",
  "Community college" = "#d7191c"
)
cat_sizes <- c(
  "R1"                = 1.4,
  "R2"                = 1.0,
  "Other 4-year"      = 0.5,
  "Community college" = 0.5
)

ggplot() +
  geom_sf(data = inst_sf, aes(color = category, size = category), alpha = 0.6) +
  scale_color_manual(values = cat_colors, name = NULL) +
  scale_size_manual( values = cat_sizes,  guide = "none") +
  labs(title    = "Postsecondary institutions in the continental US",
       subtitle = paste(nrow(institutions), "institutions · HD2022 · ICLEVEL 1–2 only")) +
  theme_void() +
  theme(legend.position = "bottom")
```

```{r save-institutions}
saveRDS(institutions, here("data", "institutions", "institutions.rds"))
cat("Saved institutions.rds:", nrow(institutions), "rows\n")
```

---

## Part 3: ZIP Code Distance Pipeline

For each continental-US ZIP code, we compute the Haversine distance (km) to
every institution and derive:

- `dist_inst_km` — distance to the single nearest institution (primary)
- Per-category nearest distance — `dist_r1_km`, `dist_r2_km`,
  `dist_other4_km`, `dist_cc_km`
- Counts within 50 km — overall and per category

ZIP-code centroids come from `zipcodeR::zip_code_db`. We restrict to the
continental US (latitude 24°–50°, longitude −125° to −66°) to match the
participant sample.

```{r zip-centroids}
zips <- zip_code_db |>
  filter(
    !is.na(lat), !is.na(lng),
    lat  > 24 & lat  < 50,
    lng > -125 & lng < -66
  ) |>
  select(zip_code = zipcode, zip_lat = lat, zip_lon = lng)

cat("ZIP codes in continental US:", format(nrow(zips), big.mark = ","), "\n")
```

```{r distance-matrix}
# Full ZIP × institution distance matrix (km). With ~33k ZIPs × ~4.4k
# institutions, this is ~145M cells (~1.1 GB) — heavy but manageable.
# distm() returns a numeric matrix; we divide by 1000 to convert m → km.
dist_mat <- distm(
  cbind(zips$zip_lon, zips$zip_lat),
  cbind(institutions$lon, institutions$lat),
  fun = distHaversine
) / 1000

cat("Distance matrix:", nrow(dist_mat), "ZIPs ×",
    ncol(dist_mat), "institutions\n")
```

```{r derive-measures}
# Category indicators (logical vectors over institution columns)
is_r1     <- institutions$category == "R1"
is_r2     <- institutions$category == "R2"
is_other4 <- institutions$category == "Other 4-year"
is_cc     <- institutions$category == "Community college"

# Nearest institution overall — index, name, category, distance
nearest_idx <- max.col(-dist_mat, ties.method = "first")
dist_inst   <- dist_mat[cbind(seq_len(nrow(dist_mat)), nearest_idx)]

nearest_in_subset <- function(cols_lgl) {
  apply(dist_mat[, cols_lgl, drop = FALSE], 1, min)
}

n_within <- function(cols_lgl, radius_km = 50) {
  rowSums(dist_mat[, cols_lgl, drop = FALSE] < radius_km)
}

zipcode_distances <- zips |>
  mutate(
    dist_inst_km        = dist_inst,
    nearest_institution = institutions$name[nearest_idx],
    nearest_category    = institutions$category[nearest_idx],
    dist_r1_km          = nearest_in_subset(is_r1),
    dist_r2_km          = nearest_in_subset(is_r2),
    dist_other4_km      = nearest_in_subset(is_other4),
    dist_cc_km          = nearest_in_subset(is_cc),
    n_within_50km          = rowSums(dist_mat < 50),
    n_r1_within_50km       = n_within(is_r1),
    n_r2_within_50km       = n_within(is_r2),
    n_other4_within_50km   = n_within(is_other4),
    n_cc_within_50km       = n_within(is_cc)
  )

# Free the big matrix
rm(dist_mat); invisible(gc())

head(zipcode_distances)
```

### Validation

```{r validation}
zipcode_distances |>
  select(dist_inst_km, dist_r1_km, dist_r2_km, dist_other4_km, dist_cc_km) |>
  summary()
```

```{r spot-checks}
# NYC (10001) → near 0 to many; rural Montana (59931) → large
zipcode_distances |>
  filter(zip_code %in% c("10001", "59931", "90210", "60601", "73301")) |>
  select(zip_code, nearest_institution, nearest_category,
         dist_inst_km, dist_r1_km, dist_cc_km, n_within_50km) |>
  mutate(across(starts_with("dist_"), \(x) round(x, 1))) |>
  knitr::kable(caption = "Spot checks: selected ZIP codes")
```

```{r dist-distribution}
#| fig-width: 7
#| fig-height: 4

zipcode_distances |>
  select(dist_inst_km, dist_r1_km, dist_r2_km, dist_other4_km, dist_cc_km) |>
  pivot_longer(everything(), names_to = "measure", values_to = "km") |>
  mutate(measure = factor(measure,
    levels = c("dist_inst_km", "dist_r1_km", "dist_r2_km",
               "dist_other4_km", "dist_cc_km"),
    labels = c("Nearest (any)", "Nearest R1", "Nearest R2",
               "Nearest other 4-yr", "Nearest community college"))) |>
  ggplot(aes(x = km)) +
  geom_histogram(bins = 60, fill = "#2c7bb6", color = "white", linewidth = 0.2) +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~measure, scales = "free_y") +
  labs(
    x        = "Distance (km, log scale)",
    y        = "ZIP codes",
    title    = "Distance to nearest institution, by category",
    subtitle = "Right-skewed → log-transform used in analysis"
  ) +
  theme_minimal(base_size = 11)
```

```{r save-distances}
saveRDS(zipcode_distances, here("data", "institutions", "zipcode_distances.rds"))
cat("Saved zipcode_distances.rds:", nrow(zipcode_distances), "rows ·",
    ncol(zipcode_distances), "columns\n")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
