---
title: "Scientist Demographics: IPEDS and ACS Comparison"
author: "Jan Pfänder"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This document characterises the demographic composition of the US scientific workforce on
**race/ethnicity**, **gender**, and **income** across two complementary data sources:

| Source | What it measures | Race/Gender | Income |
|---|---|---|---|
| **IPEDS Completions C2022_A** | Doctoral degree pipeline — who earns a PhD | ✓ | — |
| **ACS PUMS 5-year (2018–22)** | Employed workforce — who works as a scientist | ✓ | ✓ |

Each source is analysed at three levels of scope:

- **All researchers** — the broadest reference category: anyone completing a doctoral
  research degree (IPEDS) or working in a science, social science, math/statistics, or
  postsecondary teaching occupation (ACS)
- **Natural sciences** — physical, biological, earth, and environmental sciences
- **Climate-adjacent** — atmospheric, geological/earth, oceanographic, ecological, and
  natural resource sciences; the fields most directly relevant to climate research

**Outputs** (gitignored, written alongside the raw inputs they derive from):

- `data/scientist_demographics/ipeds/ipeds_demographics.rds` — tidy IPEDS summary: `scope`, `race_eth`, `gender`, `n`, `pct`
- `data/scientist_demographics/census_pus/acs_demographics.rds` — tidy ACS summary: `scope`, `race_eth`, `gender`, `n`, `pct`
- `data/scientist_demographics/census_pus/acs_income.rds` — individual-level ACS PUMS rows for employed scientists (income analysis)

**Raw data sources** — both must exist locally before rendering this document.
Each is produced by a standalone script in `R/`; run each once and the outputs
sit in `data/scientist_demographics/` for re-use.

| Source | Script | Output |
|---|---|---|
| IPEDS C2022_A (PhD completions) | `R/download_ipeds.R` | `data/scientist_demographics/ipeds/C2022_A.csv` |
| ACS PUMS 5-year (employed workforce) | `R/download_pums.R` | `data/scientist_demographics/census_pus/pums_raw.rds` |

The PUMS download takes 1–3 hours via the Census API; do **not** re-run it
unless you must. See `R/download_pums.R` for prerequisites (free Census API
key required).

---

## Setup

```{r packages}
library(tidyverse)
library(tidycensus)   # only for the pums_variables metadata lookup
library(here)
```


```{r plot-functions}
list.files(here("R", "functions"), pattern = "\\.R$", full.names = TRUE) |>
  walk(source)
```

---

## Part 1: IPEDS Completions — Degree Pipeline

We download IPEDS Completions (C2022_A), a census of all degrees awarded at US postsecondary
institutions in 2021–22, broken down by institution, CIP code, award level, race/ethnicity,
and gender. We analyse **doctoral research degrees only** (AWLEVEL = 17), which excludes
professional doctorates (JD, MD, PharmD).

### Load and clean

Reads `data/scientist_demographics/ipeds/C2022_A.csv`. Re-generate by running
`Rscript R/download_ipeds.R` (one-time, ~30s).

```{r ipeds-load}
ipeds_csv <- here("data", "scientist_demographics", "ipeds", "C2022_A.csv")
if (!file.exists(ipeds_csv)) {
  stop("IPEDS CSV not found. Run: Rscript R/download_ipeds.R")
}
comp_raw <- read_csv(ipeds_csv, show_col_types = FALSE)
```

```{r ipeds-clean}
# Replace IPEDS suppression sentinels (-1, -2) with NA before aggregating
comp_clean <- comp_raw |>
  mutate(across(where(is.numeric), \(x) if_else(x < 0, NA_real_, as.numeric(x))))
```

### Helper: aggregate race/gender totals

```{r ipeds-helper}
# Column names for race (M = men, W = women) across all race/ethnicity categories
race_gender_cols <- c(
  "CTOTALM", "CTOTALW",
  "CAIANM",  "CAIANW",   # Native American / Alaska Native
  "CASIAM",  "CASIAW",   # Asian
  "CBKAAM",  "CBKAAW",   # Black
  "CHISPM",  "CHISPW",   # Hispanic
  "CNHPIM",  "CNHPIW",   # Native Hawaiian / Pacific Islander
  "CWHITM",  "CWHITW",   # White
  "C2MORM",  "C2MORW",   # Two or more races
  "CUNKNM",  "CUNKNW",   # Unknown
  "CNRALM",  "CNRALW"    # Non-resident alien
)

# Takes a pre-filtered IPEDS dataframe (already MAJORNUM==1, AWLEVEL==17),
# aggregates national totals, and returns a tidy race × gender tibble.
aggregate_ipeds_demo <- function(df, scope_label) {
  totals <- df |>
    summarise(across(all_of(race_gender_cols), \(x) sum(x, na.rm = TRUE)))

  tibble(
    race_eth = rep(c(
      "Native American", "Asian", "Black", "Hispanic",
      "Pacific Islander", "White", "Two or more",
      "Unknown", "Non-resident alien"
    ), each = 2),
    gender = rep(c("Men", "Women"), times = 9),
    n = c(
      totals$CAIANM, totals$CAIANW,
      totals$CASIAM, totals$CASIAW,
      totals$CBKAAM, totals$CBKAAW,
      totals$CHISPM, totals$CHISPW,
      totals$CNHPIM, totals$CNHPIW,
      totals$CWHITM, totals$CWHITW,
      totals$C2MORM, totals$C2MORW,
      totals$CUNKNM, totals$CUNKNW,
      totals$CNRALM, totals$CNRALW
    ),
    scope = scope_label
  ) |>
    mutate(pct = round(100 * n / sum(n, na.rm = TRUE), 1))
}
```

### 1a: All doctoral research degrees

No CIP restriction — every AWLEVEL = 17 degree across all fields. This is the broadest
reference: anyone earning a research doctorate at a US institution.

```{r ipeds-all}
ipeds_all <- comp_clean |>
  filter(MAJORNUM == 1, AWLEVEL == 17)

cat("CIP codes (all doctoral):", n_distinct(ipeds_all$CIPCODE), "distinct codes\n")
cat("Institutions:            ", n_distinct(ipeds_all$UNITID), "\n")

ipeds_demo_all <- aggregate_ipeds_demo(ipeds_all, "All doctoral (all fields)")
```

### 1b: Natural sciences

CIP codes included: **03** (Natural Resources & Conservation), **26** (Biological &
Biomedical Sciences), **27** (Mathematics & Statistics), **40** (Physical Sciences).

```{r ipeds-natural}
ipeds_natural <- comp_clean |>
  filter(
    str_starts(CIPCODE, "03") | str_starts(CIPCODE, "26") |
    str_starts(CIPCODE, "27") | str_starts(CIPCODE, "40"),
    MAJORNUM == 1,
    AWLEVEL  == 17
  )

cat("CIP codes (natural sciences):", n_distinct(ipeds_natural$CIPCODE), "distinct codes\n")
cat("Institutions:               ", n_distinct(ipeds_natural$UNITID), "\n")

ipeds_demo_natural <- aggregate_ipeds_demo(ipeds_natural, "Natural sciences")
```

### 1c: Climate-adjacent fields

CIP codes included: **40.04** (Atmospheric Sciences), **40.06** (Geological/Earth Sciences),
**40.07** (Oceanography), **26.13** (Ecology & Environmental Biology), **03** (Natural
Resources & Conservation).

```{r ipeds-climate}
ipeds_climate <- comp_clean |>
  filter(
    str_starts(CIPCODE, "40.04") | str_starts(CIPCODE, "40.06") |
    str_starts(CIPCODE, "40.07") | str_starts(CIPCODE, "26.13") |
    str_starts(CIPCODE, "03"),
    MAJORNUM == 1,
    AWLEVEL  == 17
  )

cat("CIP codes (climate-adjacent):", n_distinct(ipeds_climate$CIPCODE), "\n")
ipeds_climate |> pull(CIPCODE) |> unique() |> sort() |> print()

ipeds_demo_climate <- aggregate_ipeds_demo(ipeds_climate, "Climate-adjacent")
```

### 1d: Tables and plots

```{r ipeds-gender-table}
bind_rows(ipeds_demo_all, ipeds_demo_natural, ipeds_demo_climate) |>
  group_by(scope, gender) |>
  summarise(n = sum(n, na.rm = TRUE), .groups = "drop") |>
  group_by(scope) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  rename(Scope = scope, Gender = gender, N = n, `%` = pct) |>
  knitr::kable(
    caption  = "IPEDS: Gender composition of doctoral completions (2022)",
    format.args = list(big.mark = ",")
  )
```

```{r ipeds-race-table}
bind_rows(ipeds_demo_all, ipeds_demo_natural, ipeds_demo_climate) |>
  filter(!race_eth %in% c("Non-resident alien", "Unknown")) |>
  group_by(scope, race_eth) |>
  summarise(n = sum(n, na.rm = TRUE), .groups = "drop") |>
  group_by(scope) |>
  mutate(pct_domestic = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  arrange(scope, desc(n)) |>
  rename(Scope = scope, `Race/ethnicity` = race_eth, N = n, `% (domestic)` = pct_domestic) |>
  knitr::kable(
    caption = "IPEDS: Race/ethnicity of doctoral completions (2022, domestic completers only)",
    format.args = list(big.mark = ",")
  )
```

```{r ipeds-plot}
#| fig-cap: "IPEDS 2022: share of doctoral completions by race/ethnicity and gender across three scopes. Domestic completers only (non-resident alien and unknown excluded)."
#| fig-width: 10
#| fig-height: 5

bind_rows(ipeds_demo_all, ipeds_demo_natural, ipeds_demo_climate) |>
  filter(!race_eth %in% c("Non-resident alien", "Unknown")) |>
  group_by(scope) |>
  mutate(pct_domestic = round(100 * n / sum(n, na.rm = TRUE), 1)) |>
  ungroup() |>
  mutate(scope = factor(scope,
    levels = c("All doctoral (all fields)", "Natural sciences", "Climate-adjacent"))) |>
  ggplot(aes(x = pct_domestic, y = reorder(race_eth, pct_domestic), fill = gender)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(Men = plot_palette[1], Women = plot_palette[4]), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  facet_wrap(~scope) +
  labs(
    x        = "Share of doctoral completions (%)",
    y        = NULL,
    title    = "Representation in doctoral programs",
    subtitle = "IPEDS Completions C2022_A · PhD only · domestic completers"
  ) +
  plot_theme
```

---

## Part 2: ACS PUMS — Employed Workforce

We use the ACS 5-year Public Use Microdata Sample (PUMS, 2018–2022), which pools five survey
years for the largest possible sample of employed scientists. Occupation is identified using
Census detailed occupation codes stored in the `OCCP` variable.

### Occupation code discovery

```{r acs-occp-lookup}
# val_min holds the numeric occupation code; sprintf gives the 4-char string
# matching the OCCP values returned by get_pums() ("1860", "1840", etc.)
occp_labels <- pums_variables |>
  filter(year == 2022, survey == "acs5", var_code == "OCCP", !is.na(val_min)) |>
  transmute(
    val_code  = sprintf("%04d", as.integer(val_min)),
    val_label
  )

cat("All OCCP labels available in pums_variables:\n")
print(occp_labels |> arrange(val_label), n = Inf)
```

```{r acs-occp-define}
# Helper: exclude technician roles (they appear alongside scientists in some
# ACS labels and would otherwise be caught as false positives)
not_technician <- function(x) !str_detect(x, regex("technician", ignore_case = TRUE))

# ── Climate-adjacent (most specific) ──────────────────────────────────────────
# Matches: 1710 Atmospheric, 1750 Geoscientists/Hydrologists,
#          1745 Environmental Scientists, 1640 Conservation Scientists/Foresters
occp_climate_adjacent <- occp_labels |>
  filter(
    str_detect(val_label, regex(
      "atmospheric|geoscient|hydrolog|environmental scientist|conservation scientist|forester",
      ignore_case = TRUE
    )),
    not_technician(val_label)
  ) |>
  pull(val_code)

# ── Natural sciences (superset of climate-adjacent) ───────────────────────────
# Adds: 1600 Agricultural/Food Scientists, 1700 Astronomers/Physicists,
#       1610 Biological Scientists, 1720 Chemists, 1650 Other Life Scientists,
#       1760 Physical Scientists
occp_natural_science_extra <- occp_labels |>
  filter(
    str_detect(val_label, regex(
      "biolog|chemist|physicist|astronomer|agricultural.*scientist|life scientist|physical scientist",
      ignore_case = TRUE
    )),
    not_technician(val_label)
  ) |>
  pull(val_code)

occp_natural_science <- union(occp_climate_adjacent, occp_natural_science_extra)

# ── All researchers (superset of natural sciences) ────────────────────────────
# Adds: 1800 Economists, 1821/1822/1825 Psychologists,
#       1860 Other Social Scientists (covers sociologists, anthropologists,
#            political scientists — no separate ACS codes in 2022),
#       1840 Urban Planners, 2205 Postsecondary Teachers,
#       plus mathematicians/statisticians from the math occupations group
occp_all_research_extra <- occp_labels |>
  filter(str_detect(val_label, regex(
    "economist|psychologist|social scientist|mathematician|statistician|postsecondary teacher|urban.*planner",
    ignore_case = TRUE
  ))) |>
  pull(val_code)

occp_all_research <- union(occp_natural_science, occp_all_research_extra)

# ── Verification ──────────────────────────────────────────────────────────────
stopifnot(all(occp_climate_adjacent %in% occp_natural_science))
stopifnot(all(occp_natural_science  %in% occp_all_research))

cat("Climate-adjacent (", length(occp_climate_adjacent), "):\n")
occp_labels |> filter(val_code %in% occp_climate_adjacent) |> arrange(val_label) |> print(n = Inf)
cat("\nNatural sciences — additions (", length(setdiff(occp_natural_science, occp_climate_adjacent)), "):\n")
occp_labels |> filter(val_code %in% setdiff(occp_natural_science, occp_climate_adjacent)) |> arrange(val_label) |> print(n = Inf)
cat("\nAll researchers — additions (", length(setdiff(occp_all_research, occp_natural_science)), "):\n")
occp_labels |> filter(val_code %in% setdiff(occp_all_research, occp_natural_science)) |> arrange(val_label) |> print(n = Inf)
```

### Load ACS PUMS

Reads `data/scientist_demographics/pums_raw.rds` produced by `R/download_pums.R`. Re-generating that file takes 1–3 hours via the Census
API; do not re-run unless necessary.

```{r acs-load}
pums_path <- here("data", "scientist_demographics", "census_pus", "pums_raw.rds")
if (!file.exists(pums_path)) {
  stop("PUMS RDS not found. Run: Rscript R/download_pums.R (slow, 1–3 hours)")
}
pums_raw <- readRDS(pums_path)

cat("PUMS rows loaded:", format(nrow(pums_raw), big.mark = ","), "\n")
```

```{r acs-recode}
# Keep employed civilians only (ESR 1 = at work, 2 = has job but not at work)
# Recode race/ethnicity and gender to match IPEDS categories
pums_sci <- pums_raw |>
  filter(ESR %in% c("1", "2")) |>
  mutate(
    gender   = if_else(SEX == "1", "Men", "Women"),
    race_eth = case_when(
      HISP != "01"              ~ "Hispanic",
      RAC1P == "1"              ~ "White",
      RAC1P == "2"              ~ "Black",
      RAC1P == "6"              ~ "Asian",
      RAC1P %in% c("3","4","5") ~ "Native American",
      RAC1P == "7"              ~ "Pacific Islander",
      RAC1P == "9"              ~ "Two or more",
      TRUE                      ~ "Unknown"
    ),
    wage   = if_else(as.numeric(WAGP) > 0, as.numeric(WAGP), NA_real_),
    weight = as.numeric(PWGTP)
  )
```

### Helper: aggregate race/gender totals

```{r acs-helper}
# Takes a pre-filtered ACS tibble, returns weighted race × gender summary
aggregate_acs_demo <- function(df, scope_label) {
  df |>
    group_by(race_eth, gender) |>
    summarise(n = sum(weight), .groups = "drop") |>
    mutate(
      pct   = round(100 * n / sum(n), 1),
      scope = scope_label
    )
}
```

### 2a: All research/academic occupations

Includes all natural and social scientists, mathematicians, statisticians, and postsecondary
teachers. This is the broadest employed-workforce parallel to the IPEDS all-doctoral scope.

```{r acs-all}
acs_all <- pums_sci |> filter(OCCP %in% occp_all_research)

cat("Rows (all research occupations, employed):", format(nrow(acs_all), big.mark = ","), "\n")
cat("By occupation (top 20 by weighted N):\n")
acs_all |>
  left_join(occp_labels, by = c("OCCP" = "val_code")) |>
  count(OCCP, val_label, wt = weight) |>
  arrange(desc(n)) |>
  slice_head(n = 20) |>
  print()

acs_demo_all <- aggregate_acs_demo(acs_all, "All research occupations")
```

### 2b: Natural science occupations

```{r acs-natural}
acs_natural <- pums_sci |> filter(OCCP %in% occp_natural_science)

cat("Rows (natural sciences, employed):", format(nrow(acs_natural), big.mark = ","), "\n")

acs_demo_natural <- aggregate_acs_demo(acs_natural, "Natural sciences")
```

### 2c: Climate-adjacent occupations

```{r acs-climate}
acs_climate <- pums_sci |> filter(OCCP %in% occp_climate_adjacent)

cat("Rows (climate-adjacent, employed):", format(nrow(acs_climate), big.mark = ","), "\n")
cat("By occupation:\n")
acs_climate |>
  left_join(occp_labels, by = c("OCCP" = "val_code")) |>
  count(OCCP, val_label, wt = weight) |>
  arrange(desc(n)) |>
  print()

acs_demo_climate <- aggregate_acs_demo(acs_climate, "Climate-adjacent")
```

### 2d: Tables and plots

```{r acs-gender-table}
bind_rows(acs_demo_all, acs_demo_natural, acs_demo_climate) |>
  group_by(scope, gender) |>
  summarise(n = sum(n), .groups = "drop") |>
  group_by(scope) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  rename(Scope = scope, Gender = gender, N = n, `%` = pct) |>
  knitr::kable(
    caption = "ACS: Gender composition of employed scientists (2018–22, weighted)",
    format.args = list(big.mark = ",")
  )
```

```{r acs-race-table}
bind_rows(acs_demo_all, acs_demo_natural, acs_demo_climate) |>
  filter(race_eth != "Unknown") |>
  group_by(scope) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  arrange(scope, desc(n)) |>
  rename(Scope = scope, `Race/ethnicity` = race_eth, N = n, `%` = pct) |>
  knitr::kable(
    caption = "ACS: Race/ethnicity of employed scientists (2018–22, weighted, unknown excluded)",
    format.args = list(big.mark = ",")
  )
```

```{r acs-demo-plot}
#| fig-cap: "ACS 2018–22: share of employed scientists by race/ethnicity and gender across three scopes. Weighted estimates; unknown race excluded."
#| fig-width: 10
#| fig-height: 5

bind_rows(acs_demo_all, acs_demo_natural, acs_demo_climate) |>
  filter(race_eth != "Unknown") |>
  group_by(scope) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  ungroup() |>
  mutate(scope = factor(scope,
    levels = c("All research occupations", "Natural sciences", "Climate-adjacent"))) |>
  ggplot(aes(x = pct, y = reorder(race_eth, pct), fill = gender)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c(Men = plot_palette[1], Women = plot_palette[4]), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  facet_wrap(~scope) +
  labs(
    x        = "Share of employed workers (%)",
    y        = NULL,
    title    = "Representation among employed researchers",
    subtitle = "ACS PUMS 2018–22 · employed civilians · weighted"
  ) +
  plot_theme
```

### 2e: Income

```{r acs-income-table}
bind_rows(
  mutate(acs_all,     scope = "All research occupations"),
  mutate(acs_natural, scope = "Natural sciences"),
  mutate(acs_climate, scope = "Climate-adjacent")
) |>
  filter(!is.na(wage)) |>
  group_by(scope) |>
  summarise(
    n              = sum(weight),
    median_wage    = round(matrixStats::weightedMedian(wage, weight) / 1000) * 1000,
    pct_over_100k  = round(
      100 * sum((wage > 100000) * weight) / sum(weight), 1
    ),
    pct_over_150k  = round(
      100 * sum((wage > 150000) * weight) / sum(weight), 1
    ),
    .groups = "drop"
  ) |>
  mutate(scope = factor(scope,
    levels = c("All research occupations", "Natural sciences", "Climate-adjacent"))) |>
  arrange(scope) |>
  rename(
    Scope = scope, N = n,
    `Median wage` = median_wage,
    `% >$100k` = pct_over_100k,
    `% >$150k` = pct_over_150k
  ) |>
  knitr::kable(
    caption = "ACS: Income of employed scientists (2018–22, weighted)",
    format.args = list(big.mark = ",")
  )
```

```{r acs-income-plot}
#| fig-cap: "Income distribution by scope (ACS PUMS 2018–22, weighted). Dashed line = US median earnings for all workers 16+ (~$48,060 in 2022)."
#| fig-width: 7
#| fig-height: 4

bind_rows(
  mutate(acs_all,     scope = "All research occupations"),
  mutate(acs_natural, scope = "Natural sciences"),
  mutate(acs_climate, scope = "Climate-adjacent")
) |>
  filter(!is.na(wage)) |>
  mutate(scope = factor(scope,
    levels = c("All research occupations", "Natural sciences", "Climate-adjacent"))) |>
  ggplot(aes(x = wage / 1000, color = scope, weight = weight)) +
  geom_density(linewidth = 0.8, adjust = 1.5) +
  geom_vline(xintercept = 48.06, linetype = "dashed",
             color = "grey40", linewidth = 0.4) +
  annotate("text", x = 51, y = Inf, vjust = 1.5, hjust = 0,
           label = "US median", size = 3, color = "grey40") +
  scale_x_continuous(
    limits = c(0, 300),
    labels = scales::dollar_format(suffix = "k", scale = 1)
  ) +
  scale_color_manual(
    values = plot_palette[c(1, 3, 4)],
    name   = NULL
  ) +
  labs(
    x        = "Annual wage / salary income",
    y        = "Density",
    title    = "Income distribution by scope",
    subtitle = "ACS PUMS 2018–22 · employed civilians · weighted"
  ) +
  plot_theme
```

---

## Part 3: Cross-Source Comparison (Climate-Adjacent)

IPEDS measures the PhD degree pipeline (who earns a doctorate); ACS measures the employed
workforce (who works as a scientist). Differences between the two highlight barriers beyond
degree completion — in hiring, retention, or career progression.

```{r compare-prep}
# Align race/ethnicity labels; drop Non-resident alien and Unknown for valid comparison
ipeds_for_compare <- ipeds_demo_climate |>
  filter(!race_eth %in% c("Non-resident alien", "Unknown")) |>
  group_by(race_eth, gender) |>
  summarise(n = sum(n, na.rm = TRUE), .groups = "drop") |>
  group_by(gender) |>
  mutate(
    pct    = round(100 * n / sum(n), 1),
    source = "IPEDS (PhD pipeline, 2022)"
  ) |>
  ungroup()

acs_for_compare <- acs_demo_climate |>
  filter(race_eth != "Unknown") |>
  group_by(race_eth, gender) |>
  summarise(n = sum(n), .groups = "drop") |>
  group_by(gender) |>
  mutate(
    pct    = round(100 * n / sum(n), 1),
    source = "ACS (employed workforce, 2018–22)"
  ) |>
  ungroup()
```

```{r compare-table}
bind_rows(ipeds_for_compare, acs_for_compare) |>
  select(source, race_eth, gender, pct) |>
  pivot_wider(names_from = source, values_from = pct) |>
  arrange(race_eth, gender) |>
  knitr::kable(
    caption = "Pipeline vs. workforce: race/ethnicity by gender (climate-adjacent fields, % within gender group)",
    digits  = 1
  )
```

```{r compare-plot}
#| fig-cap: "PhD pipeline (IPEDS 2022) vs. employed workforce (ACS 2018–22) for climate-adjacent science fields. Differences show where demographic gaps widen or narrow between degree completion and employment."
#| fig-width: 8
#| fig-height: 5

bind_rows(ipeds_for_compare, acs_for_compare) |>
  ggplot(aes(x = pct, y = reorder(race_eth, pct),
             color = source, shape = source)) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(plot_palette[1], plot_palette[4]),
    name   = NULL
  ) +
  scale_shape_manual(values = c(16, 17), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.1))) +
  facet_wrap(~gender) +
  labs(
    x        = "Share within gender group (%)",
    y        = NULL,
    title    = "PhD pipeline vs. employed workforce: climate-adjacent science",
    subtitle = "IPEDS 2022 · ACS 2018–22 · domestic completers"
  ) +
  plot_theme +
  theme(legend.position = "top")
```

---

## Save Outputs

```{r save-outputs}
# Combine IPEDS results (all three scopes)
ipeds_demographics <- bind_rows(ipeds_demo_all, ipeds_demo_natural, ipeds_demo_climate)
saveRDS(ipeds_demographics, here("data", "scientist_demographics", "ipeds", "ipeds_demographics.rds"))
cat("Saved ipeds/ipeds_demographics.rds:", nrow(ipeds_demographics), "rows\n")

# Combine ACS demographics results (all three scopes)
acs_demographics <- bind_rows(acs_demo_all, acs_demo_natural, acs_demo_climate)
saveRDS(acs_demographics, here("data", "scientist_demographics", "census_pus", "acs_demographics.rds"))
cat("Saved census_pus/acs_demographics.rds:", nrow(acs_demographics), "rows\n")

# ACS individual rows for income analysis (all three scopes)
saveRDS(
  bind_rows(
    mutate(acs_all,     scope = "All research occupations"),
    mutate(acs_natural, scope = "Natural sciences"),
    mutate(acs_climate, scope = "Climate-adjacent")
  ),
  here("data", "scientist_demographics", "census_pus", "acs_income.rds")
)
cat("Saved census_pus/acs_income.rds\n")
```

---

## Session Info

```{r session-info}
sessionInfo()
```
